---
title: "BCB420 - A1"
output:
  html_document:
    df_print: paged
author: "Mit Patel"
---

Setup and package installation: 
```{r setup, message=FALSE, results='hide'}
knitr::opts_chunk$set(warning = FALSE)
install.packages(pkgs = c("BiocManager","dplyr"),
         repos = "http://cran.rstudio.org",
         dependencies = TRUE,
         quiet = TRUE)
library(BiocManager)
library(Biobase)
library(GEOquery)
```

## Downloading dataset and supplementary files 
```{r data download, message=FALSE, results='hide'}
gset <- getGEO("GSE108539", GSEMatrix=FALSE, getGPL=FALSE)
current_gpl <- names(GPLList(gset))
current_gpl_info <- Meta(getGEO(current_gpl))


```

## Information about Platform
**Platform Title:** `r current_gpl_info$title`\
**Original submission date:** `r current_gpl_info$submission_date`\
**Last update date:** `r current_gpl_info$last_update_date`\
**Organism:** `r current_gpl_info$organism`\
**No. of GEO datasets that use this technology:** `r length(current_gpl_info$series_id)`\
**No. of GEO samples that use this technology:** `r length(current_gpl_info$sample_id)`

```{r}

sfiles = getGEOSuppFiles('GSE108539')
fnames = rownames(sfiles)
exp_data = read.delim(fnames[1],header=TRUE,
                check.names = FALSE)
```

```{r class example}
sfiles = getGEOSuppFiles('GSE70072')
fnames_class = rownames(sfiles)
ca125_exp = read.delim(fnames_class[1],header=TRUE,
                check.names = FALSE)

samples <- data.frame(lapply(colnames(ca125_exp)[3:22], 
        FUN=function(x){unlist(strsplit(x, split = "\\."))[c(2,3)]}))
colnames(samples) <- colnames(ca125_exp)[3:22]
rownames(samples) <- c("patients","cell_type")
samples <- data.frame(t(samples))

```

```{r}

```
